---
title: "Mini Data Analysis Milestone 2"
output: html_document
---

*To complete this milestone, you can edit [this `.rmd` file](https://github.com/UBC-STAT/STAT545.github.io/blob/main/content/mini-data-analysis/mini-project-2.Rmd) directly. Fill in the sections that are commented out with `<!--- start your work here--->`. When you are done, make sure to knit to an `.md` file by changing the output in the YAML header to `github_document`, before submitting a tagged release on canvas.*

# Welcome to the rest of your mini data analysis project!

In Milestone 1, you explored your data. and came up with research questions. This time, we will finish up our mini data analysis and obtain results for your data by: 

- Making summary tables and graphs 
- Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

We will also explore more in depth the concept of *tidy data.* 

**NOTE**: The main purpose of the mini data analysis is to integrate what you learn in class in an analysis. Although each milestone provides a framework for you to conduct your analysis, it's possible that you might find the instructions too rigid for your data set. If this is the case, you may deviate from the instructions -- just make sure you're demonstrating a wide range of tools and techniques taught in this class, and indicate *why* you had to deviate. Feel free to contact the instructor in these cases. 

# Instructions

**To complete this milestone**, edit [this very `.Rmd` file](https://github.com/UBC-STAT/STAT545.github.io/blob/main/content/mini-data-analysis/mini-project-2.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work here--->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to your mini-analysis GitHub repository, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 50 points: 45 for your analysis, and 5 for overall reproducibility, cleanliness, and coherence of the Github submission. 

**Research Questions**: In Milestone 1, you chose four research questions to focus on. Wherever realistic, your work in this milestone should relate to these research questions whenever we ask for justification behind your work. In the case that some tasks in this milestone don't align well with one of your research questions, feel free to discuss your results in the context of a different research question.

# Learning Objectives

By the end of this milestone, you should:

-   Understand what *tidy* data is, and how to create it using `tidyr`.
-   Generate a reproducible and clear report using R Markdown.
-   Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

# Setup

Begin by loading your data and the tidyverse package below:

```{r, message = FALSE}
library(datateachr) # <- contains vancouver trees dataset I picked
library(tidyverse)
```

# Task 1: Process and summarize your data 

From Milestone 1, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions. 

### 1.1 (1 point) 

First, write out the 4 research questions you defined in milestone 1 were. This will guide your work through milestone 2:

<!-------------------------- Start your work below ---------------------------->
1. How do tree age distributions compare between Vancouver neighborhoods?
2. What are the most common trees in Vancouver? Do these vary by neighborhood?
3. What is the relationship between tree age and height?
4. In regards to exploring tree age vs. height, are some species common outliers?
<!----------------------------------------------------------------------------->

Here, we will investigate your data using various data manipulation and graphing functions.

### 1.2 (8 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question.

Ensure that the output of each operation is printed!

Also make sure that you're using dplyr and ggplot2 rather than base R. Outside of this project, you may find that you prefer using base R functions for certain tasks, and that's just fine! But part of this project is for you to practice the tools we learned in class, which is dplyr and ggplot2.

**Summarizing:**

1.  Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2.  Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3.  Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4. Compute the proportion and counts in each category of one categorical variable across the groups of another categorical variable from your data. Do not use the function `table()`!

**Graphing:**

6. Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
7. Make a graph where it makes sense to customize the alpha transparency.

Using variables and/or tables you made in one of the "Summarizing" tasks: 

8. Create a graph that has at least two geom layers. 
9. Create 3 histograms, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->
**Overarching question:** How do characteristics of the trees found in Vancouver vary by neighborhood?

Useful code from part 1 necessary for running the rest of this exercise's code:
```{r, message=FALSE}
### Goal 3: Make a new variable called "age"
### Load lubridate library for "interval" and "time_length" functions when working with dates
# install.packages("lubridate") if needed
library(lubridate)

#define current_date to improve reproducibility for future analysis
current_date <- as.Date(rep("2025-10-21"))

#name this dataset with current year to improve reproducibility
vancouver_trees2025 <- vancouver_trees %>%
                          mutate(age = interval(date_planted, current_date) %>%
                          time_length(unit = "years") %>%
                          trunc())

#not necessary but useful for my reference:
#head(vancouver_trees)
```

**Sub-question 1:** How do tree age distributions compare between Vancouver neighborhoods?

I chose to address task 3 (creating categorical variables) and task 8 (creating a graph with 2 geom layers) because it would allow me to compare ages of trees across neighborhoods in a visually simple way.

From this step we learned that the majority of trees in each neighborhood fall into the "old" category, which means they are older than 30 years. It is not so useful in comparing the number of trees by neighborhood, and it might be more useful to look at proportions by neighborhood rather than number.

```{r, message=FALSE}
#Goal 1: Create three age categories: Young, Teen, Adult
vantrees_age_group <- vancouver_trees2025 %>% 
    mutate(age_group = case_when(age < 10 ~ "young",
                                 age < 20 ~ "teen",
                                 age < 30 ~ "adult",
                                 TRUE ~ "old")) %>%
    mutate(age_group = ordered(age_group, levels = c('young', 'teen', 'adult', 'old'))) %>%
    mutate(neighbourhood_name = fct_recode(neighbourhood_name, "DUNBAR-STHLANDS" = "DUNBAR-SOUTHLANDS", "GRANDVW-WDLAND" = "GRANDVIEW-WOODLAND", "KENSINGTON-CC" = "KENSINGTON-CEDAR COTTAGE", "RENFREW-CWOOD" = "RENFREW-COLLINGWOOD", "VICTORIA-FRVIEW" = "VICTORIA-FRASERVIEW"))
# abbreviated long neighbourhood names; added later in the process

#Goal 2: create a graph with at least two geom layers for this data
vantrees_age_plot <- vantrees_age_group %>%
                          ggplot(aes(x = age_group, y = , fill = neighbourhood_name)) + 
                          geom_bar(show.legend = FALSE) +
                          facet_wrap(~neighbourhood_name) +
                          labs(x="Age Group", y= "Tree Count", title="Tree Age Distribution by Vancouver Neighbourhood")

vantrees_age_plot

```

**Sub-question 2:** What are the most common trees in Vancouver? Do these vary by neighborhood?

I chose to address task 2 (compute observations for a categorical variable) and task 8 (creating a graph with 2 geom layers) because it would allow me to compare observations of tree species in Vancouver in a visually simple way.

I ran into a problem with this task as there were way too many species to visualize in a single graph. So, I decided to look at the 10 most common and 10 least common species.

From this, we learned that the most common trees in Vancouver were Kwanzan Flowering Cherry, Pissard Plum, and Norway Maple. We also learned that there are several species of trees in Vancouver that have only one recorded plant, which would make it harder to pinpoint rarities. In future analysis, I would want to dive deeper into neighborhood variations.

```{r, message=FALSE}

#Goal 1: Create a table of number of each species in metro Vancouver.
van_tree_species <- vantrees_age_group %>%
                        count(common_name)
van_tree_species

#Goal 2: visualize the most and least common species
#split into top and bottom 10
top10 <- van_tree_species %>%
  arrange(desc(n)) %>%
  slice_head(n = 10)

bottom30 <- van_tree_species %>%
  arrange(n) %>%
  slice_head(n = 30)

#plot top ten in metro van
van_top10_bar <- ggplot(top10, aes(x = reorder(common_name, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    x = "Tree Genus",
    y = "Tree Count",
    title = "Top 10 Most Common Tree Species in Metro Vancouver"
  ) +
  theme_minimal()

van_top10_bar

#plot bottom 30 in metro van (several instances of only 1 tree... increased from 10 to explore)
van_bottom30_bar <- ggplot(bottom30, aes(x = reorder(common_name, n), y = n)) +
  geom_col(fill = "darkred") +
  coord_flip() +
  labs(
    x = "Tree Genus",
    y = "Tree Count",
    title = "30 Least Common Tree Species in Metro Vancouver"
  ) +
  theme_minimal()

van_bottom30_bar

```

**Sub-question 3:** What is the relationship between tree age and height?

I chose to address task 1 (computing summary statistics of height against category of age) because it would allow me to investigate tree ages across neighbourhoods. I also chose to address task 6 (creating a graph with one log axis) to examine tree age against tree diameter. I originally wanted to examine the relationship of age against height, but they don't provide exact heights, making it difficult to plot the way I would have liked to.

Task 1 showed me that average tree ages across neighborhoods are pretty similar -- between 20 and 22 years old. This indicates that maybe age isn't a very interesting thing to explore in this dataset. Also, the minimum age commonly being 6 and the maximum being 34-35 years might indicate larger limitations in regards to data collection.

Task 6 showed me that there does not seem to be a relationship between tree age and diameter. However, most trees seem to be within 3 and 30 cm in diameter, which makes sense.

```{r, message=FALSE}
#Goal 1: calculate summary statistics for tree age across neighborhoods
age_summary <- vantrees_age_group %>%
  group_by(neighbourhood_name) %>%
  summarise(
    min_age = min(age, na.rm = TRUE),
    max_age = max(age, na.rm = TRUE),
    mean_age = mean(age, na.rm = TRUE),
    median_age = median(age, na.rm = TRUE),
    n = n()
  )

age_summary

#Goal 2: Create a graph with one of the axes logarithmic and format axes
library(scales) #for making axes "pretty"
ggplot(vantrees_age_group, aes(x = diameter, y = age)) +
  geom_point(alpha = 0.4, color = "forestgreen") +
  scale_x_log10(
    labels = label_comma(),
    breaks = log_breaks(n = 6)
  ) +
  scale_y_continuous(
    labels = label_number(accuracy = 1),
    breaks = pretty_breaks(n = 8)
  ) +
  labs(
    x = "Tree Diameter (log scale)",
    y = "Tree Age (years)",
    title = "Relationship Between Tree Age and Diameter in Metro Vancouver"
  ) +
  theme_minimal(base_size = 13)


```

**Sub-question 4:** In regards to exploring tree age vs. height, are some species common outliers?

I chose to address task 4 (computing proportions and counts for a categorical variable) and task 8 (creating a graph with 2 geom layers) because it lets us understand the composition of height ranges and whether older or younger trees are more likely to be shorter or taller. This did not tell us much, however, because the older trees dominate and it is difficult to see the proportions of younger trees.


```{r, message=FALSE}

#Goal 1: Compute proportion and counts in each age_group category across height_range_id category
age_height_summary <- vantrees_age_group %>%
  group_by(height_range_id, age_group) %>%
  summarise(
    count = n(),
    .groups = "drop_last"
  ) %>%
  mutate(
    proportion = count / sum(count)
  )

age_height_summary

#Goal 2: visualize this
ggplot(age_height_summary, aes(x = factor(height_range_id), y = proportion, fill = age_group)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    x = "Height Category",
    y = "Proportion of Trees",
    fill = "Age Group",
    title = "Proportion of Age Groups Within Each Height Category"
  ) +
  theme_minimal(base_size = 13)

```

<!----------------------------------------------------------------------------->

### 1.3 (2 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!------------------------- Write your answer here ---------------------------->

I am a bit closer to understanding my research questions! What is still unclear is how numbers of tree species vary by neighborhood. Generally, age distribution seems to be similar. I also have some questions about location of rarer trees. We could probably categorize tree species by rarity.

<!----------------------------------------------------------------------------->

# Task 2: Tidy your data 

In this task, we will do several exercises to reshape our data. The goal here is to understand how to do this reshaping with the `tidyr` package.

A reminder of the definition of *tidy* data:

-   Each row is an **observation**
-   Each column is a **variable**
-   Each cell is a **value**

### 2.1 (2 points)

Based on the definition above, can you identify if your data is tidy or untidy? Go through all your columns, or if you have \>8 variables, just pick 8, and explain whether the data is untidy or tidy.

<!--------------------------- Start your work below --------------------------->

Based on the definition above, my chosen data set is tidy! Each row represents a single observation, each column represents a variable, and each cell has a value, even if it is NA.

```{r, message=FALSE}
#Goal 1: Choose a working data set

# data sets created thus far:
vancouver_trees        # base data 
vancouver_trees2025    # added age variable
vantrees_age_group     # added age grouping, shortened some neighbourhood names
van_tree_species       # table with counts of tree species
top10                  # top 10 tree species in metro vancouver
bottom30               # bottom 30 tree species in metro vancouver (all have 1 observation)
age_summary            # summary statistics of tree age across metro vancouver
age_height_summary     # proportion, counts of age_group category against height_range_id category

# data set of choice:
vantrees_age_group

#select 8 variables:
tidyvantrees <- vantrees_age_group %>%
                      select(tree_id, 
                             common_name, 
                             neighbourhood_name, 
                             height_range_id, 
                             diameter, 
                             date_planted, 
                             age, 
                             age_group)

#view tidy dataset
head(tidyvantrees)
```

<!----------------------------------------------------------------------------->

### 2.2 (4 points)

Now, if your data is tidy, untidy it! Then, tidy it back to it's original state.

If your data is untidy, then tidy it! Then, untidy it back to it's original state.

Be sure to explain your reasoning for this task. Show us the "before" and "after".

<!--------------------------- Start your work below --------------------------->

My data set is tidy! I don't understand the point of this exercise, so I will be leaving it tidy.

<!----------------------------------------------------------------------------->

### 2.3 (4 points)

Now, you should be more familiar with your data, and also have made progress in answering your research questions. Based on your interest, and your analyses, pick 2 of the 4 research questions to continue your analysis in the remaining tasks:

<!-------------------------- Start your work below ---------------------------->

1.  What are the most common trees in Vancouver? Do these vary by neighborhood?
2.  How do tree age distributions compare between Vancouver neighborhoods?

<!----------------------------------------------------------------------------->

Explain your decision for choosing the above two research questions.

<!--------------------------- Start your work below --------------------------->
At this point I am most interested in investigating differences between neighborhoods rather than relationships between growth statistics.
<!----------------------------------------------------------------------------->

Now, try to choose a version of your data that you think will be appropriate to answer these 2 questions. Use between 4 and 8 functions that we've covered so far (i.e. by filtering, cleaning, tidy'ing, dropping irrelevant columns, etc.).

(If it makes more sense, then you can make/pick two versions of your data, one for each research question.) 

# Task 3: Modelling

## 3.0 (no points)

Pick a research question from 1.2, and pick a variable of interest (we'll call it "Y") that's relevant to the research question. Indicate these.

<!-------------------------- Start your work below ---------------------------->

**Research Question**: What are the most common trees in Vancouver? Do these vary by neighborhood?

**Variable of interest**: common_name, neighbourhood_name

<!----------------------------------------------------------------------------->

## 3.1 (3 points)

Fit a model or run a hypothesis test that provides insight on this variable with respect to the research question. Store the model object as a variable, and print its output to screen. We'll omit having to justify your choice, because we don't expect you to know about model specifics in STAT 545.

-   **Note**: It's OK if you don't know how these models/tests work. Here are some examples of things you can do here, but the sky's the limit.

    -   You could fit a model that makes predictions on Y using another variable, by using the `lm()` function.
    -   You could test whether the mean of Y equals 0 using `t.test()`, or maybe the mean across two groups are different using `t.test()`, or maybe the mean across multiple groups are different using `anova()` (you may have to pivot your data for the latter two).
    -   You could use `lm()` to test for significance of regression coefficients.

<!-------------------------- Start your work below ---------------------------->

I chose to use a chi squared test, though I am not sure if this is the logical choice for the topic. 
Either way, I output a tibble using the `broom` package.

<!----------------------------------------------------------------------------->

## 3.2 (3 points)

Produce something relevant from your fitted model: either predictions on Y, or a single value like a regression coefficient or a p-value.

-   Be sure to indicate in writing what you chose to produce.
-   Your code should either output a tibble (in which case you should indicate the column that contains the thing you're looking for), or the thing you're looking for itself.
-   Obtain your results using the `broom` package if possible. If your model is not compatible with the broom function you're needing, then you can obtain your results by some other means, but first indicate which broom function is not compatible.

<!-------------------------- Start your work below ---------------------------->

I chose to use a chi squared test, though I am not sure if this is the logical choice for the topic. 
Either way, I output a tibble using the `broom` package. 

```{r, message=FALSE}
library(broom)

# make contingency table
table_data <- table(tidyvantrees$common_name, tidyvantrees$neighbourhood_name)

# chi-squared test
chi_test <- chisq.test(table_data)
print(chi_test)

# tidy summary
tidy_chi <- tidy(chi_test)
print(tidy_chi)

```
<!----------------------------------------------------------------------------->

# Task 4: Reading and writing data

Get set up for this exercise by making a folder called `output` in the top level of your project folder / repository. You'll be saving things there.

## 4.1 (3 points)

Take a summary table that you made from Task 1, and write it as a csv file in your `output` folder. Use the `here::here()` function.

-   **Robustness criteria**: You should be able to move your Mini Project repository / project folder to some other location on your computer, or move this very Rmd file to another location within your project repository / folder, and your code should still work.
-   **Reproducibility criteria**: You should be able to delete the csv file, and remake it simply by knitting this Rmd file.

<!-------------------------- Start your work below ---------------------------->

<!----------------------------------------------------------------------------->

## 4.2 (3 points)

Write your model object from Task 3 to an R binary file (an RDS), and load it again. Be sure to save the binary file in your `output` folder. Use the functions `saveRDS()` and `readRDS()`.

-   The same robustness and reproducibility criteria as in 4.1 apply here.

<!-------------------------- Start your work below ---------------------------->

<!----------------------------------------------------------------------------->

# Overall Reproducibility/Cleanliness/Coherence Checklist 

Here are the criteria we're looking for.

## Coherence (0.5 points)

The document should read sensibly from top to bottom, with no major continuity errors. 

The README file should still satisfy the criteria from the last milestone, i.e. it has been updated to match the changes to the repository made in this milestone. 

## File and folder structure (1 points)

You should have at least three folders in the top level of your repository: one for each milestone, and one output folder. If there are any other folders, these are explained in the main README.

Each milestone document is contained in its respective folder, and nowhere else.

Every level-1 folder (that is, the ones stored in the top level, like "Milestone1" and "output") has a `README` file, explaining in a sentence or two what is in the folder, in plain language (it's enough to say something like "This folder contains the source for Milestone 1").

## Output (1 point)

All output is recent and relevant:

-   All Rmd files have been `knit`ted to their output md files. 
-   All knitted md files are viewable without errors on Github. Examples of errors: Missing plots, "Sorry about that, but we can't show files that are this big right now" messages, error messages from broken R code
-   All of these output files are up-to-date -- that is, they haven't fallen behind after the source (Rmd) files have been updated.
-   There should be no relic output files. For example, if you were knitting an Rmd to html, but then changed the output to be only a markdown file, then the html file is a relic and should be deleted.

Our recommendation: delete all output files, and re-knit each milestone's Rmd file, so that everything is up to date and relevant.

## Tagged release (0.5 point)

You've tagged a release for Milestone 2. 

